---
title: "p8105_hw2_lk2761"

author: "Lorraine Kwok"

date: "09/27/2019"

output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
  )

library(tidyverse)
library(readxl)
```

## Problem 1

This problem uses the 2019 Mr. Trash Wheel dataset focusing on the data for Mr. Trash wheel and precipitation in 2017 and 2018. 

```{r clean Mr. Trash Wheel sheet}
# import and clean Mr. Trash Wheel sheet 

trash_wheel_data = 
  read_excel("data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", 
             sheet = 1, 
             range = cell_cols("A:N")) %>%
  janitor::clean_names(dat = .) %>%
  drop_na(data = ., dumpster) %>%
  mutate(.data = ., 
         sports_balls = round(sports_balls, digits = 0),
         sports_balls_int = as.integer(sports_balls),
         homes_powered_rd = round(homes_powered, digits = 2)) %>%
  select(.data = ., -sports_balls, -homes_powered)

trash_wheel_data
```

```{r df for precipitation 2017 and 2018}
# import and clean 2017 precipitation data 

precip_2017 = 
  read_excel("data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", 
             sheet = 6) %>%
  janitor::clean_names(dat = .) %>%
  rename(.data = ., month = precipitation_in, total_inches = x2) %>%
  filter(.data = ., month != "Month") %>%
  drop_na(data = .)  %>%
  mutate(.data = ., year = 2017) 

precip_2017

# import and clean 2018 precipitation data

precip_2018 = 
   read_excel("data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", 
             sheet = 5) %>%
  janitor::clean_names(dat = .) %>%
  rename(.data = ., month = precipitation_in, total_inches = x2) %>%
  filter(.data = ., month != "Month") %>%
  drop_na(data = .) %>%
  mutate(.data = ., year = 2018) 

precip_2018

# create new dataset by combining precip_2017 and precip_2018

precip_combined = 
  bind_rows(precip_2017, precip_2018) %>%
  mutate(.data = ., 
         total_in = round(as.numeric(total_inches), digits = 2),
         month = month.name[as.numeric(month)]) %>% 
  select(.data = ., year, month, total_in, -total_inches)

precip_combined
```

The Mr. Trash Wheel dataset consists of `r nrow(trash_wheel_data)` observations. The mean weight of trash removed by Mr. Trash Wheel in the Inner Harbor of Baltimore, Maryland, as denoted by the variable *weight_tons* between May 2014 and June 2019 is `r round(mean(pull(trash_wheel_data, weight_tons)), 2)` tons and the mean volume, denoted by *volume_cubic_yards* is `r round(mean(pull(trash_wheel_data, volume_cubic_yards)), 2)` cubic yards. The median number of sports balls removed in 2017 was `r median(pull(trash_wheel_data, sports_balls_int))` balls. 

Data on precipitation in 2017 and 2018 were combined into one dataset, which is made up of `r nrow(precip_combined)` observations. The average amount of rainfall from January 2017 to December 2018 was `r round(mean(pull(precip_combined, total_in)), 2)` inches. The total precipitation in 2018 was `r sum()

## Problem 2

We are now using the FiveThirtyEight data.

```{r pols_month dataframe}
# create dataframe with data on politicians
# In this dataset, for prez_gop, there were values of "2," which still resulted in coding of "0" in the prez_dem variable. 

pols_month_df =
  read.csv("./data/fivethirtyeight_datasets/pols-month.csv") %>%
  janitor::clean_names(dat = .) %>%
  separate(., mon, into = c("year", "month", "day")) %>%
  mutate(.data = .,
         prez_dem = recode(prez_dem, `1` = "dem", `0` = "gop"), 
         president = prez_dem,
         month = month.name[as.numeric(month)]) %>%
  select(.data = ., -prez_dem, -prez_gop, -day)

pols_month_df
```

```{r snp dataframe}
# import and clean S&P closing data 

snp_df =
  read.csv("./data/fivethirtyeight_datasets/snp.csv") %>%
  janitor::clean_names(dat = .) %>%
  rename(.data = ., closing_values = close) %>%
  separate(., date, into = c("month", "day", "year")) %>% 
  select(.data = ., year, everything(), -day) %>%
  mutate(.data = ., month = month.name[as.numeric(month)])

snp_df
```

```{r unemployment dataframe}
# import and clean unemployment data

unemploy_df =
  read.csv("./data/fivethirtyeight_datasets/unemployment.csv") %>%
  janitor::clean_names(dat = .) 

unemploy_df
# convert the months from rows to columns in the unemployment data frame

unemploy_data_col = 
  pivot_longer(
    unemploy_df,
    jan:dec,
    names_to = "month",
    values_to = "percent_unemploy")

unemploy_data_col

# convert the abbreviated months to full names 

unemploy_tidy_data = 
  unemploy_data_col %>%
  mutate(.data = ., 
         month = recode(month, 
                        `jan` = "January", 
                        `feb` = "February",
                        `mar` = "March", 
                        `apr` = "April",
                        `may` = "May",
                        `jun` = "June",
                        `jul` = "July",
                        `aug` = "August",
                        `sep` = "September",
                       `oct` = "October",
                        `nov` = "November",
                        `dec` = "December"))

unemploy_tidy_data
```

The following code chunk combines the three individual datasets, seen above, together.

```{r merge datasets}
# merge pols_month_df and snp_df 

pol_snp_data =
  left_join(snp_df, pols_month_df, by = c("year", "month")) %>%
  mutate(.data = ., year = as.integer(year))

# merge unemploy_tidy_data into pols_snp_data 

pol_snp_unemploy_data =
  left_join(unemploy_tidy_data, pol_snp_data, by = c("year", "month"))
```

As a result of merging three separate datasets, the resulting dataset is `r nrow(pol_snp_unemploy_data)` rows by `r ncol(pol_snp_unemploy_data)` columns. The data are for the years, 1948-2015. This dataset consists of key variables around whether  national politicians (governors, senators, presidents and representatives) were Democratic or Republican at a given year (*gov_gop*, *gov_dem*, *sen_gop*, *sen_dem*, *rep_gop*, *rep_dem*, *president*), the closing values of the Standard & Poor's (S&P) stock market index in a given month and year (*closing_values*), and the percentage of unemployment in a given month and year (*percent_unemploy*). 

## Problem 3

We are using the NYC Popular Baby Names dataset. 

```{r baby_names dataframe}
# import Popular_Baby_Names csv and clean data 

baby_names_df = 
  read.csv("./data/Popular_Baby_Names.csv") %>%
  janitor::clean_names(dat = .) %>%
  rename(.data = ., child_first_name = child_s_first_name, num_children = count) %>%
  mutate(.data = ., 
         child_first_name = str_to_upper(child_first_name),
         ethnicity = recode(ethnicity, 
                            `ASIAN AND PACI` = "ASIAN AND PACIFIC ISLANDER",
                            `BLACK NON HISP` = "BLACK NON HISPANIC",
                            `WHITE NON HISP` = "WHITE NON HISPANIC")) %>%
  distinct(.data = ., .keep_all = FALSE)
```

We will create a reader-friendly table showing the rank in popularity of the name "Olivia" as a female baby name over time. 

```{r create table for Olivia}
baby_names_df %>%
  filter(.data = ., child_first_name == "OLIVIA") %>%
  pivot_wider(names_from = "year_of_birth",
              values_from = "rank") %>%
  select(.data = ., -child_first_name, everything(), -num_children, -gender) %>%
  group_by(.data = ., ethnicity) %>%
  fill("2016", "2015", "2014", "2013", "2012", "2011") %>%
  fill("2016", "2015", "2014", "2013", "2012", "2011", .direction = 'up') %>%
  distinct() %>%
  knitr::kable(caption = "Table 1. Popularity of the name, Olivia, by ethnicity and year")
```

We will create a reader-friendly table showing the most popular male name (rank = 1) for all ethnicities and all years. 

```{r create table for popular male baby names}
baby_names_df %>%
  filter(.data = ., gender == "MALE", rank == "1") %>%
  pivot_wider(names_from = "year_of_birth",
              values_from = "child_first_name") %>%
  select(.data = ., -rank, -num_children, -gender) %>%
  group_by(.data = ., ethnicity) %>%
  fill("2016", "2015", "2014", "2013", "2012", "2011") %>%
  fill("2016", "2015", "2014", "2013", "2012", "2011", .direction = 'up') %>%
  distinct() %>%
  knitr::kable(caption = "Table 2. Most popular male first name by ethnicity and year")
```

We will now create a scatterplot to display the data on the popularity of first names for male, white non-hispanic children born in 2016

```{r create scatterplot}
baby_names_df %>%
  filter(.data = ., 
         year_of_birth == "2016" & 
         gender == "MALE" & 
         ethnicity == "WHITE NON HISPANIC") %>%
  ggplot(aes(x = rank, y = num_children)) +
  geom_point(aes(), alpha = .5) + 
  labs(
    title = "Popularity of first name among male, white non-hispanic children born in 2016",
    x = "Rank in popularity of name",
    y = "Number of children with name"
  ) +
  theme_minimal()
```

